A Computational Account of Default Mode Function\\ by Control Theory and Reinforcement Learning

### 2017-03-28 paper/paper.tex
fixup
fixups
model-free RL is plausible in the brain via dopamine

In artificial intelligence (AI), a popular abstract model for
multi-step decision processes in such an environment are so-called
\textit{Partially Observable Markov Decision Processes (POMDPs)}.
A POMDP models a decision process in which it is assumed that environment dynamics are determined by a Markov process,
but the agent cannot directly observe the underlying state. Instead, it tries to optimize a \textit{subjective} reward
signal which is tried alongside each transition in the environments state-space, and maintains a probability distribution over the set of possible
states, based on a set of observations and observation probabilities. This is a minimal amount of assumptions that can be made about an environment,
and is characteristic of so-called \textit{model-free RL}, the approach adapted by our work.
In contrast, \textit{model-based RL} starts off with some mechanistic assumptions about the dynamics of world.
These can be assumptions about the physical laws governing the agent's world or constraints on the state-space and transitions thereupon.
For example if a rat in a maze knows, for sure,
that staying still will produce no change in the environment, and in particular will not eventually lead to finding the food.
% A core property of human intelligence has been proposed underly to improve
% on an expected utility outcomes as a strategy for action choice in uncertain
% environments \citep{gershman2015computational}.
Our rat might represent such knowledge about the world as:
\begin{itemize}
\item $r(s, \text{``stand still''}) = 0$ if $s$ does not correspond to a cell / chamber containing food.
\item $p(s'|s,\text{``stand still''}) = 1$ if $s'=s$ and $0$ otherwise.
\item etc.
\end{itemize}

@@

\subsection{Conclusion: putting everything together}
\begin{figure}[!h]
  \includegraphics[width=.9\linewidth]{rl_process_chart.pdf}
  \caption{The DMN as an RL agent...}
  \label{fig:rl_process_chart}
\end{figure}
The DMN is today known to consistently increase in neural
  activity when engaging cognitive processes that are detached from
  current sensory environment \citep{kenet2003spontaneously, fiser2004small}.
  Its involvement in thinking about the past,
  the future, and hypothetical possibilities ties in with the implicit computation of
  action and state cascades as a function of what happened in the past.
% A policy represents the repertoire of possible actions
%   on the world given a current state. It encodes the probabilites of
%   choosing actions to be executed in a certain situation.
% Interactions between external and internal sources of
%   information are probably commonplace during perception \citep{gershman2010learning}
The DMN may subserve
  constant exploration of possible future actions and their
  cumulative outcomes. Implicit computation of future choices
  provides an explanation for purposeful mind-wandering.

@@

% \begin{figure}[!tb]
%   \centering
%   \includegraphics[width=.8\linewidth]{vae.pdf}
%   \caption{Variational autoencoders...}
%   \label{fig:vae}
% \end{figure}
The main algorithm for minimizing free-energy $F_G^R(\x)$ is the \textit{wake-sleep algorithm}.
\citep{dayan1995helmholtz}. As noted already in \citep{dayan1995helmholtz},
a crucial drawback of the wake-sleep algorithm is that it requires a concurrent models (generative and recogntiion), which together do not correspond to optimization of (a bound of) the marginal likelihood (because of the incorrect KL used therein, etc.). Thus the brain could not possible be running such a algorithm, not even in principle! To the rescue, we note that the recent theory of
\textit{variational auto-encoders (VAEs)} \citep{kingma2013auto} might provide an efficient alternative to the wake-sleep algorithm, as it overcomes the technical limits of the former, by using a reparametrization trick. For example, unlike the wake-sleep algorithm for minimizing free-energy, VAEs can be efficiently trained back-propagation of learning errors.

>>>>

In artificial intelligence (AI), a popular abstract model for
multi-step decision processes in such an environment are so-called
\textit{Partially Observable Markov Decision Processes (POMDPs)}.
A POMDP models a decision process in which it is assumed that environment dynamics are determined by a Markov process,
but the agent cannot directly observe the underlying state. Instead, it tries to optimize a \textit{subjective} reward
signal which is tried alongside each transition in the environments state-space, and maintains a probability distribution over the set of possible
states, based on a set of observations and observation probabilities. This is a minimal amount of assumptions that can be made about an environment,
and is characteristic of so-called \textit{model-free RL}, the approach adapted by our work.
\begin{mdframed}
\textbf{Model-free RL in the brain is plausible.}  Indeed, as was noted in \cite{dayan2008decision}, It has long been suggested in the literature that there
  is a rather direct
mapping of model-free RL learning algorithms
onto the brain, with the neuromodulator dopamine serving
as a teaching signal to train values or policies by controlling
synaptic plasticity at targets such as the ventral and
dorsolateral striatum.
\end{mdframed}

In contrast, \textit{model-based RL} starts off with some mechanistic assumptions about the dynamics of world.
These can be assumptions about the physical laws governing the agent's world or constraints on the state-space and transitions thereupon.
For example if a rat in a maze knows, for sure,
that staying still will produce no change in the environment, and in particular will not eventually lead to finding the food.
% A core property of human intelligence has been proposed underly to improve
% on an expected utility outcomes as a strategy for action choice in uncertain
% environments \citep{gershman2015computational}.
Our rat might represent such knowledge about the world as:
\begin{itemize}
\item $r(s, \text{``stand still''}) = 0$ if $s$ does not correspond to a cell / chamber containing food.
\item $p(s'|s,\text{``stand still''}) = 1$ if $s'=s$ and $0$ otherwise.
\item etc.
\end{itemize}

@@

\subsection{Putting everything together}
\begin{figure}[!h]
  \includegraphics[width=.9\linewidth]{rl_process_chart.pdf}
  \caption{The DMN as an RL agent...}
  \label{fig:rl_process_chart}
\end{figure}
The DMN is today known to consistently increase in neural
  activity when engaging cognitive processes that are detached from
  current sensory environment \citep{kenet2003spontaneously, fiser2004small}.
  Its involvement in thinking about the past,
  the future, and hypothetical possibilities ties in with the implicit computation of
  action and state cascades as a function of what happened in the past.
% A policy represents the repertoire of possible actions
%   on the world given a current state. It encodes the probabilites of
%   choosing actions to be executed in a certain situation.
% Interactions between external and internal sources of
%   information are probably commonplace during perception \citep{gershman2010learning}
The DMN may subserve
  constant exploration of possible future actions and their
  cumulative outcomes. Implicit computation of future choices
  provides an explanation for purposeful mind-wandering.

@@

% \begin{figure}[!tb]
%   \centering
%   \includegraphics[width=.8\linewidth]{vae.pdf}
%   \caption{Variational autoencoders...}
%   \label{fig:vae}
% \end{figure}
The main algorithm for minimizing free-energy $F_G^R(\x)$ is the \textit{wake-sleep algorithm}.
\citep{dayan1995helmholtz}. As noted already in \citep{dayan1995helmholtz},
a crucial drawback of the wake-sleep algorithm is that it requires a concurrent models (generative and recogntiion), which together do not correspond to optimization of (a bound of) the marginal likelihood (because of the incorrect KL used therein, etc.). Thus the brain could not possible be running such a algorithm, not even in principle! To the rescue, we note that the recent theory of
\textit{variational auto-encoders (VAEs)} \citep{kingma2013auto} might provide an efficient alternative to the wake-sleep algorithm, as it overcomes the technical limits of the former, by using a reparametrization trick. For example, unlike the wake-sleep algorithm for minimizing free-energy, VAEs can be efficiently trained via back-propagation of learning errors.
